---
title: "Assignment 4C: RUNNING AND INTERPRETING MDS ANALYSES"
author: "Jung"
date: "February 8, 2021"
output: html_document
---

## PART 1:
### 1) Using isoMDS in R, run nonmetric (ordinal) MDS analyses of the data from Howe on similarity of crimes, in 1-4 dimensions.  See the posted R example file for guidance.
 
```{r}

# load libraries
library(MASS)

# read in the data file
data.sim <- read.csv("CRIMES_PRX_clean.csv",header=TRUE)

# store the rownames/colnames into names
names <- data.sim[,1]

# convert to matrix
data.sim <- data.matrix(data.sim)

# remove the first column and make it as a row names
rownames(data.sim)<-names
data.sim <- data.sim[,-1]

# find maximum value 
my_max=max(data.sim, na.rm=TRUE)
my_max

```
```{r}
# dissimilarity matrix
data.dis = 190 - data.sim
data.dis

```
```{r}

# need to make the matrix symmetric
data.dis[is.na(data.dis)] <- 0
data.dis = (data.dis+t(data.dis))
data.dis

```
```{r}

dimstress <- matrix(numeric(8),4,2)

#dimension 1
dimstress[1,1] <-1
dimstress[1,2] <- isoMDS(data.dis,k=1,trace=TRUE,p=2)$stress

#dimension 2
dimstress[2,1] <-2
dimstress[2,2] <- isoMDS(data.dis,k=2,trace=TRUE,p=2)$stress

#dimension 3
dimstress[3,1] <-3
dimstress[3,2] <- isoMDS(data.dis,k=3,trace=TRUE,p=2)$stress

#dimension 4
dimstress[4,1] <-4
dimstress[4,2] <- isoMDS(data.dis,k=4,trace=TRUE,p=2)$stress

dimstress
```
### 2) Prepare a plot of Stress vs. dimensionality for these four solutions. Judging from this plot, what do you think is the best dimensionality for these data? Why?

#### Answer: Dimensionality = 4 seems to be the best dimensionality for the data, as it has the lowest value of sterss.


```{R}

plot(dimstress, xlab = "Dimensionality", ylab = "Stress")

```

### 3) Plot and label the 2-dim. configuration (regardless of your conclusions about dimensionality in part 2).  Report your (subjective) interpretation for this solution.


```{r}

data.nmds <- isoMDS(data.dis,k=2,trace=TRUE,p=2)
data.nmds

longdata <- data.nmds$points
longdata <- as.data.frame(longdata)
longdata$names <- rownames(data.nmds$points)
longdata
library(ggplot2)
plot1 <- ggplot(longdata, aes(V1, V2, label=names)) + 
  geom_point(colour="red", size=2) +
  geom_text(colour="red", check_overlap = TRUE, size=2.5, 
            hjust = "left", vjust = "bottom", nudge_x = 0.02, nudge_y = 0) + 
  labs(x="", y="", title="MDS by isoMDS()") + theme_bw()
plot1

```

## PART 2:
### 4) As an aid to interpretation: Using the 2-dimensional configuration you obtain from isoMDS, use a statistical package (R or SPSS) to regress the attribute of SERIOUSNESS into the solution space.

```{r}

#drop the third column, mnames
longdata = subset(longdata, select = -c(3) )
longdata

serious <- c(7.83,6.08,8.5,2.17,6.92,6.42,8.33,4.67,7.96,8.96,6.46,7.46,8.71,5.21,6.71,6.83,4.63,5.5)
serious <- as.data.frame(serious)

#bind "seriousness"
df <- cbind(longdata,serious)
df

colnames(df) <- c("x1","x2","Y")
```

### Plot (by hand, or using the stat package) the regression line into a plot of the configuration space (one ignores the constant term in doing this).  

```{r}

#Regression
fit<-lm(Y ~ x1 + x2, data=df)
summary(fit)

#coefficient for x1
c1 <- fit$coefficients[2]

#coefficient for x2
c2 <- fit$coefficients[3]

plot2 <- plot1 +
   geom_segment(data=fit, mapping=aes(x=0, y=0, xend=c1, yend=c2), size=1, arrow=arrow(),color="green")
plot2
```

