---
title: "smacof"
author: "Chloe Jung"
date: "2/11/2021"
output: html_document
---
## MDS of the new nations data using smacof

As part of a classroom demo in a previous semester, we collected similarity ratings of 12 nations, a semi-replication of Wish's nations data collected in the 1970's. It is not a perfect replication, because at least one of the countries Wish used as stimuli (Yugoslavia) no longer exists. The file "nations_ALLDATA_1920_lowerhalf.csv" contains the averaged similarity ratings from two years sections of this course, as a lowerhalf matrix.

#### 1) read the new nations data into R, and convert it to type matrix (using the first column as the rownames)

```{r}

# load packages
library(readr)
library(tidyverse)
library(smacof)
library(ggplot2)

# import the file
nnations <-read_csv("nations_ALLDATA_1920_lowerhalf.csv")

# convert to matrix
nnations<- nnations %>% remove_rownames %>% column_to_rownames(var="X1")
nn <- as.matrix(nnations)

# dissmilarity matrix
Dx <- sim2diss(nn,method=8) #since the max value in nn is 7, chose to subtract from 8

```

#### 2) Next, we will try using the smacof routine on the matrix Dx:
##### Run solutions in 1-5 dimensions, and choose what you consider to be the best dimensionality. Use both a “scree” plot and interpretability as your criteria for deciding on dimensionality. Write interpretations for the dimensions of your chosen solution.

```{r}

# stress for each dimension 1-5
stress_values<- matrix(numeric(10),5,2)
for(ii in 1:5){
            stress_values[ii,1] <-ii
            nn_sm <- smacofSym(Dx, ndim = ii,
                                        type =   "ordinal")
            stress_values[ii,2] <- nn_sm$stress
}
stress_values <- as.data.frame(stress_values)
colnames(stress_values) <- c("dimension","stress")

# scree plot
screeplot <-ggplot(stress_values) +
  aes(x = dimension, y = stress) +
  geom_point() +
  geom_line() +
  theme_classic()
screeplot

```

Stress is monotonically decreasing with increasing dimensionality.The elbow of the scree plot seems to be around 3 (a point where stress decreases only marginally with increasing dimensionality). However, considering the interpretability, I have chosen 2 as the dimension.

```{r}

# ndim=2
nn_sm2<-smacofSym(Dx,ndim=2, type="ordinal")
nn_sm2

# plot the 2-D solution and label the points plot
plot(nn_sm2$conf,asp=1,pch=' ') 
text(nn_sm2$conf,rownames(nn))

```

#### 3) Perform 10 runs of smacof (ordinal MDS) on the new nations data with a RANDOM start, using the init="random" parameter. Do this with 1-dimensional and 2-dimensional solutions. Do you see any evidence of a local minimum problem in any of the runs?

```{r}

# 1-dimensional solutions
fitstress<- matrix(numeric(30),10,3)
fit1 <- list()
for(i in 1:10){
            fit1[[i]] <- smacofSym(Dx, ndim = 1, type ="ordinal", init="random")
            fitstress[i,1] <-i
            fitstress[i,2] <- fit1[[i]]$stress}

# 2-dimensional solutions
fit2 <- list()
for(i in 1:10){
            fit2[[i]] <- smacofSym(Dx, ndim = 2, type ="ordinal", init="random")
            fitstress[i,3]<- fit2[[i]]$stress}

fitstress <- as.data.frame(fitstress)
colnames(fitstress) <- c("iteration","dimension1","dimension2")
fitstress

# 10 iterations plot for dimension 1
iterationplot1 <-ggplot(fitstress) +
  aes(x = iteration, y = dimension1) +
  geom_point() +
  geom_line() +
  theme_classic()
iterationplot1

# 10 iterations plot for dimension 2
iterationplot2 <-ggplot(fitstress) +
  aes(x = iteration, y = dimension2) +
  geom_point() +
  geom_line() +
  theme_classic()
iterationplot2