---
title: "PCA_JUNG"
author: "Chloe Jung"
date: "January 21, 2021"
output: html_document
---

#### Import Files 

```{r}

library(haven)

fatality <- read_sav("fatality.sav")
fatality

features <- c("state", "frate", "drink", "gasuse", "drivers", "density", "temp")

```

Features: 
- state: name of state
- frate: fatalities per 100M vehicle miles driven
- drink: min drinking age
- gasuse: gasoline use
- drivers: total number of drivers (M)
- density: population density
- temp: mean January daytime low temp

"State" will be excluded 

```{r}

# If "state" were to be included, must be converted into numeric
#fatality$state = as.numeric(as.factor(fatality$state))

```

### PCA Using "prcomp"

prcomp: The calculation is done by a singular value decomposition of the (centered and possibly scaled) data matrix, not by using eigen on the covariance matrix. This is generally the preferred method for numerical accuracy.

```{r}

fatality.pca <- prcomp(fatality[2:7], center = TRUE, scale = TRUE)
summary(fatality.pca)

```
- Standard Deviation = eigenvalues (since the data has been centered and scaled - standardized)

- Proportion of Variance = amount of variance the component accounts for in the data

- Cumulative Proportion = accumulated amount of explained variance 

### Scree Plot and Cumulative Variance Plot

```{r}

screeplot(fatality.pca, type = "l", main = "Screeplot of the PCs")
abline(h = 1, col="red", lty=5)
legend("topright", legend=c("Eigenvalue = 1"),
       col=c("red"), lty=5, cex=0.6)

# Cumulative Variance Plot
cumpro <- cumsum(fatality.pca$sdev^2 / sum(fatality.pca$sdev^2))
plot(cumpro[], xlab = "PC #", ylab = "Amount of explained variance", main = "Cumulative variance plot")
abline(v = 3, col="blue", lty=5)
abline(h =  0.8390, col="blue", lty=5)
legend("topleft", legend=c("Cut-off @ PC3"),
       col=c("blue"), lty=5, cex=0.6)

```

### Explanation for Scree Plot and Cumulative Variance Plot ###

An eigenvalues < 1 means that component explains less than a single explanatory variable, and therfore, PCs that have eigenvalues less than 1 can be discarded. 

Including PC1, PC2, and PC3 accounts for around 83.9% of the cumulative variance, which seems to be good.

### Plot PC1 and PC2 

```{r}

library(ggfortify)
pca.plot <- autoplot(fatality.pca, data = fatality, color = 'state')
pca.plot

```

### Dotplot

```{r}

# Dotplot for PC1
load <- fatality.pca$rotation
sorted.loadings <- load[order(load[, 1]), 1]
sorted.loadings
dotchart(sorted.loadings, main="Loadings Plot for PC1", xlab="Variable Loadings", cex=1, col="red")

# Dotplot for PC2
load <- fatality.pca$rotation
sorted.loadings <- load[order(load[, 2]), 2]
sorted.loadings
dotchart(sorted.loadings, main="Loadings Plot for PC2", xlab="Variable Loadings", cex=1, col="red")
```

For PC1, drivers and gasuse, as well as density and temperature, seem to have a strong loadings
For PC2, dominant variables are frate, temp, drink, gasuse and drivers.


### Biplot

```{r}

biplot(fatality.pca, cex=c(0.5, 0.7))

```


#### Varimax
- change of coordinates that maximizes the sum of variances of the squared loadings
- goal to clean up the rotation

```{r}

fatality.var = varimax(fatality.pca$rotation)
fatality.var

````

## Alternative Method


### PCA Using "princomp" on Correlation Matrix 

### Convert to correlation matrix

```{r}

# Correlation matrix
fatality.cor = cor(fatality[2:7])
fatality.cor = as.matrix(fatality.cor)
fatality.cor

```

The correlation matrix suggest that there is a strong, positive correlation between the gasoline usage (gasuse) and total number of drivers (drivers), which seems to be intuitive. There is a moderate negative correlation between the fatalities rate (frate) and population density (density), and also


### PCA and Scree Plot
```{r}

# PCA 
fatality.pca2 <- princomp(fatality[2:7], cor = TRUE, scores = TRUE, covmat = fatality.cor)
print(fatality.pca2)    ## prints out eigenvalues

# Scree Plot
plot(fatality.pca2,type='l')   ## scree plot
abline(h = 1, col="red", lty=5)
legend("topright", legend=c("Eigenvalue = 1"),
       col=c("red"), lty=5, cex=0.6)

```
The scree plot seems to suggest the same result as seen when used "prcomp," which is expected. 

### Plot PC1 and PC2

```{r}

# Plot of first two components
P12<-fatality.pca2$loadings[,1:2]  ## prepare to plot the first two components, with full variable names
plot(P12,pch=' ')  # plot components with blanks at first
text(P12, rownames(fatality.cor), cex=0.8)  # then add full subtest names

```

### Eigendecomposition

- use "eigen" directly, and define the principal components in terms of the eigendecomposition

- accomplish a PCA via the eigendecomposition routines in R function eigen returns eigenvalues (in $values) and eigenvectors (in $vectors)

```{r}

fatality.eigen<-eigen(fatality.cor,symmetric=TRUE)
fatality.eigen

fatality.eigen$values
fatality.eigen$vectors

```

# Compuing Principal Components from Eigenvectors & Eigenvalues

```{r}
# prepare to compute principal components from eigenvectors & eigenvalues

# 1. create 2x2 diagonal identity matrix
I2<-matrix(c(1,0,0,1),nrow=2,byrow=TRUE)

# 2. plug sqrt(eigenvalues) into diagonal of identity matrix to create weighting matrix "wgt"
wgt<-I2
for (i in 1:2) {wgt[i,i]<-(fatality.eigen$values[i])^.5}
print(wgt)

# 3. weight the first two eigenvectors to obtain (the first 2) principal components
P2<-fatality.eigen$vectors[,1:2] %*% wgt

# 4. plot the first two principal components
rownames(P2)<-rownames(fatality.cor)
plot(P2,pch=' ')
text(P2,rownames(P2),cex=0.8)  # add full variable names

```
### Varimax

```{r}

# most common method for orthogonal analytic rotation: VARIMAX
vmax<-varimax(P2,normalize=TRUE,eps=1e-5)
P2_vmax<-vmax$loadings
P2_vmax
plot(P2_vmax,pch=' ',asp=1,)
abline(h = 0, v = 0)
text(P2_vmax,rownames(P2),asp=1, cex=0.5)  # add full variable names

```

### Interpretation from the analysis above ##

fatality rate (frate) and population density (density) seem to have contrasting relationship, while the gas use (gasuse) and number of drivers (driver) seem to have a close relationship.

drinking age (drink) and mean temperature in January for the state (temp) seem to be highly related to the fatality rate.

States 30, 39, 21, 7 seem to have contrasting relationship with states 31, 28, 24.
